---
title: "Sherryx_OriginalHomeworkCode_01"
output: html_document
date: "2025-02-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#took out the stinger package because otherwise it would not let me knit into HTML
#THIS IS WHERE IT WOULD NOT LET ME KNIT. IT CRASHES. IT IS TELLING ME TO UPDATE MY PACKAGE WHICH I HAVE BUT IT STILL DOES NOT WORK...
#remove.packages("tidyverse")
#install.packages("tidyverse")
install.packages("stringr")
library(stringr)

```

# Matrices and Arrays

```{r}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
m
```

```{r}
m <- matrix(data = c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3, byrow = FALSE)
m
```

```{r}
m <- matrix(data = c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3, byrow = TRUE)
m
```

```{r}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2)
m1
```

```{r}
m2 <- cbind(v1, v2)
m2
```

```{r}
class(m1)
dim(m1)
class(m2)
dim(m2)
colnames(m2)
rownames(m2)
str(m2) #this is really helpful for getting overall info at one glance
```

# Challenge 1

```{r}
# Create a vector of the first line of Ulysses
line <- "Stately, plump Buck Mulligan came from the stairhead, bearing a bowl of lather on which a mirror and a razor lay crossed."

# Ambitious act of removing punctuation
line_clean <- gsub("[[:punct:]]", "", line)

# easily divide a single string into a vector of substrings = a vector of words
words <- str_split(line_clean, " ")[[1]]  

# Extract every third element
result <- words[seq(3, length(words), by = 3)] 
result

#CM: Interesting way you did this! I used the str_split() function but did not add the [[1]]. Could you explain what that was for? The problem I ran into was that if I only used str_split(), the output was a list variable which I then needed to turn into a vector using the unlist() function.
```

# Challenge 2

```{r}
m <- matrix(data = 1:80, nrow = 8, ncol = 10, byrow = FALSE)
m
```

## Given the matrix, **m**, above, extract the 2nd, 3rd, and 6th columns and assign them to the variable **x**

```{r}
x <- m[, c(2, 3, 6)] 
x

#CM: Cool! I didn't know youy could do this without giving the columns and rows their own "names." Now I know that I can do that! Could you maybe add a simple comment stating the syntax used in matrix coordinate calling? Such as matrix[row,column]. I think it can help people understand the syntax of the next couple of code chunks.
```

## Given the matrix, **m**, above, extract the 6th to 8th row and assign them to the variable **x**

```{r}
x <- m[6:8, ]
x
```

## Given the matrix, **m**, above, extract the elements from row 2, column 2 to row 6, column 9 and assign them to the variable **x**

```{r}
x <- m[2:6, 2:9]
x
```

# Challenge 3

## Construct a 4-dimensional, 400 element array (5 x 5 x 4 x 4) named **a** consisting of the numbers 400 to 1 (i.e., a descending series)

```{r}
a <- array(data = 400:1, dim = c(5, 5, 4, 4))
a
```

## Given this matrix, what would the following return?

-   a[1, 1, 1, 2]

```{r}
a[1, 1, 1, 2]
#row 1, column 1, in table 1,2 = 300

a[2, 3, 2, ]
# 364, 264, 164,  64 = row 2, cloumn 3, in dimension 3 with a 2. there are multiple values because there are only 3 instructions

a[1:5, 1:5, 3, 3]
#basically the entire table 3,3. with rows 1-5 and columns 1-5 it is the entire 3,3 table

#CM: I think having these in different code chunks would be easier to visualize and help understanding how it works piece by piece. This would be specially benficial for the last one since it does get more complex and the results/output with a more simple explanation could make it easier to understand! Right now the explanations are a bit unclear and hard to visualize
```

# Overwriting

```{r}
m[7, 1] <- 564
m
#row 7 column 1 which used to be 7 got replaced by 564
```

```{r}
m[, 8] <- 2
m
#the entire 8th column became 2s
```

```{r}
m[2:5, 4:8] <- 1
m
#like a square in the middle of the table. values from both column 4-8 and row 2-5 are substituted with 1
```

```{r}
m[2:5, 4:8] <- c(20, 19, 18, 17)
m
#row 2 from cloumn 4-8 all got substituted with 20, row 3 from cloumn 4-8 all got substituted with 19, row 4 from cloumn 4-8 all got substituted with 18, lastly row 5 from cloumn 4-8 all got substituted with 17. It went in descedning order. 
```

```{r}
m[2:5, 4:8] <- matrix(data = c(20:1), nrow = 4, ncol = 5, byrow = TRUE)
m
#going from row 2 column 4 all the way to row 5 column 8 with the vertical bounds as column 4 and 8, the values got substituted by 20 to 1 going in descending order. 
```

```{r}
m[, 8] <- c("a", "b")
m
#column 8 got substituted by alternation "a" and "b" going down the list. 
```

# Lists and Data Frames

```{r}
s <- c("this", "is", "a", "vector", "of", "strings")  # here we have a vector
m <- matrix(data = 1:40, nrow = 5, ncol = 8)  # this is a matrix
b <- FALSE  #and this is a logical statement
l <- list(s, m, b)  # this puts them all together as a list
l
```

You can extract elements from a list similarly to how you would from other data structures, except that you use **double brackets** to reference a single element in the list.

```{r}
l[[2]]  #let's take another look at our matrix
# a bit confused here... I do not see a change
#10 seconds later, figured it out. it only shows list 2 
```

An extension of this notation can be used to access elements contained within an element in the list. For example:

```{r}
l[[2]][2, 6]  #this will show us the entry at row 2, column 6 of the matrix in the second position of the list
```

To reference/extract multiple elements from a list, you would use **single bracket** notation, which would itself return a list. This is called “list slicing”.

```{r}
l[1:2]  # here we're slicing our list so it just has the first and second elements
```

```{r}
l[c(1, 3)]  # and here we're only taking the first and third elements

```

Using `str()` provides details about the three elements in our list:

```{r}
str(l)
```

You can **name** the elements in a list using the `names()` function, which adds a name attribute to each list item.

```{r}
names(l) <- c("string", "matrix", "logical")
names(l)
```

Now, you can use the name of an item in the list to refer to it:

```{r}
l$string  # this is calling the element named 'string' in list 'l'
```

```{r}
l$matrix[3, 5]  # this is calling the entry in row 3, column 5 of the element named 'matrix' in list 'l'

```

# **CHALLENGE 4**

Create a **list** representing this (simplified) primate taxonomy:

-   Primates

    -   Haplorhini

        -   Anthropoidea

            -   Platyrrhini

                -   Cebidae

                -   Atelidae

                -   Pitheciidae

            -   Catarrhini

                -   Cercopithecidae

                -   Hylobatidae

                -   Hominidae

        -   Tarsioidea

            -   Tarsiidae

    -   Strepsirhini

        -   Lorisoidea

            -   Lorisidae

            -   Galagidae

        -   Lemuroidea

            -   Cheirogaleidae

            -   Lepilemuridae

            -   Indriidae

            -   Lemuridae

            -   Daubentoniidae

```{r}
#Professor Schmit's hint code
Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae",
    "Daubentoniidae")
Lorisoidea <- c("Lorisidae", "Galagidae")
Strepsirhini <- list(Lemuroidea, Lorisoidea)
names(Strepsirhini) <- list("Lemuroidea", "Lorisoidea")
Strepsirhini

#my code trying to organize haplorhini

Tarsioidea <- c("Tarsiidae")
Platyrrhini <- c("Cebidae" , "Atelidae" , "Pitheciidae")
Catarrhini <- c("Cercopithecidae" , "Hylobatidae" , "Hominidae")
Anthropoidea <- list(Platyrrhini, Catarrhini)
names(Anthropoidea) <- list("Platyrrhini", "Catarrhini")
Anthropoidea
Haplorhini <- list(Anthropoidea, Tarsioidea)
names(Haplorhini) <- list("Anthropoidea", "Tarsioidea")
Haplorhini

#my code for primates in total
Primates <- list(Haplorhini, Strepsirhini)
names(Primates) <- list("Haplorhini" , "Strepsirhini")
Primates 

#CM: This code definitely achieves the correct structure of the primate taxonomy we're aiming to recreate. However, the code seems a bit messy and it is unclear the order in which you're approaching this. It makes maintenance easier, and ease of read for newcomers to the challenge solution. You can solve this by define lower-level groups first, then use them in higher-level lists. Again, code is correct! Just some issues with organization.
```

A **data frame** is perhaps the most useful (and most familiar) data structure that we can operate with in ***R*** as it most closely aligns with how we tend to represent tabular data, with rows as *cases* or *observations* and columns as *variables* describing those observations (e.g., a measurement of a particular type). Variables tend to be measured using the same units and thus fall into the same data class and can be thought of as analogous to vectors.

The `data.frame()` command can be used to create dataframes from scratch, like this dataframe representing data on some (fictional) BU undergrads:

```{r}
df <- data.frame(firstName = c("Nitin", "Silvy", "Ingrid", "Claire"), major = c("Biology (CM)",
    "Anthropology (Bio)", "Anthropology (Medical)", "Biology (EEB)"), sex = c("M",
    "F", "F", "F"), yearInProgram = c(1, 2, 2, 3))
df

```

Rather than creating one from scratch, we more commonly read tabular data into ***R*** from a pre-existing data file (i.e., a spreadsheet), which typically results in the table being represented as a dataframe. The following code will read from the file “random-people.csv”, currently stored on a user’s Desktop:

```{r}
df <- read.csv("//Users/sherryxie/CODE/Github/repos/AN588-Homework1-sherryx/random-people.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
# only print select columns
df[, c(1, 3, 4, 11, 12)]
```

```{r}
str(df)
```

As for other data structures, you can select and subset data frames using **bracket notation**. You can also select named columns from a data frame the **\$** operator.

```{r}
df[, 3]

```

```{r}
str(df[, 3])
# returns a vector of data in column 3
```

```{r}
df$name.last
```

```{r}
str(df$name.last)
# returns a vector of data for column *name.last*

```

```{r}
df[3]

```

```{r}
str(df[3])
# returns a data frame of data from column 3

```

```{r}
df["name.last"]

```

```{r}
str(df["name.last"])
# returns a data frame of data for column *name.last*

```

You can add rows (additional cases) or columns (additional variables) to a data frame using `rbind()` and `cbind()`.

```{r}
df <- cbind(df, id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16,
    17, 18, 19, 20))
df <- cbind(df, school = c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College",
    "MIT", "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU",
    "Harvard", "MIT", "Harvard", "BU", "UMASS Boston", "Harvard"))
df

```

Alternatively, you can extend a data frame by adding a new variable directly using the **\$** operator, like this:

```{r}
df$school <- c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College", "MIT",
    "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU", "Harvard",
    "MIT", "Harvard", "BU", "UMASS Boston", "Harvard")

```

NOTE: `cbind()` results in **school** being added as a *factor* while using the **\$** operator results in **school** being added as a *character* vector. You can see this by using the `str()` command.

```{r}
str(df)
```

# Filtering

Logical vectors may also be used to subset data frames. Here, we subset the data frame for only those rows where the variable **school** is “BU”.

```{r}
new_df <- df[df$school == "BU", ]
new_df
```

In this case, ***R*** evaluates the expression \`df\$school == “BU” and returns a logical vector equal in length to the number of rows in **df**. It then subsets **df** based on that vector, returning only rows that evaluate to **TRUE**.

We can also choose to only return particular columns when we subset.

```{r}
new_df <- df[df$school == "BU", c("name.last", "name.first", "school")]
new_df

```

# Factors, Conversion and Coercion, and Special Data Values

We were introduced to the **factor** data class above. Again, factors are numeric codes that ***R*** can use internally that correspond to character value “levels”.

When we load in data from an external source (as we will do in our next module), ***R*** tends to import character string data as factors, assigning to each unique string an integer numeric code, and assigning the string as a “label” for that code. Using factors can make some code run much more quickly (e.g., in ANOVA, ANCOVA, and other forms of regression using categorical variables).

You can convert *factor* to *character* data (and vice versa) using the `as.character()` or `as.factor()` commands. You can also convert/coerce any vector to a different class using similar constructs (e.g., `as.numeric()`), although not all such conversions are really meaningful. Converting *factor* data to *numeric* results in the the converted data having the value of ***R***’s internal numeric code for the factor level, while converting *character* data to *numeric* results in the data being coerced into the special data value of `NA` (see below) for missing data.

Finally, ***R*** has three special data values that it uses in a variety of situations.

-   `NA` (for *not available*) is used for missing data. Many statistical functions offer the possibility to include as an argument `na.rm=TRUE` (“remove NAs”) so that `NA`s are excluded from a calculation.

-   `Inf` (and `-Inf`) is used when the result of a numerical calculation is too extreme for ***R*** to express

-   `NaN` (for *not a number*) is used when ***R*** cannot express the results of a calculation , e.g., when you try to take the square root of a negative number

# **CHALLENGE 5**

## Store the following numbers as a 5 x 3 matrix, called “m3”: 3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2. Be sure to fill the matrix ROW-WISE.

```{r}
m3 <- matrix(c(3, 0, 1, 23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2), nrow = 5, ncol = 3, byrow = TRUE)

str(m3)
```

Then, do the following:

## Coerce the matrix to a data frame.

```{r}
df <- as.data.frame(m3)

str(df)
```

## As a data frame, coerce the second column to be *logical-valued*

```{r}
df[, 2] <- as.logical(df[, 2])

str(df)

# CM: Interesting approach on how you call the data! I definitely did not thing that the df[,2] would work! What I did was to call them by the provisional column names X1, X2, X3, etc.
```

## As a data frame, coerce the third column to be *factor-valued*

```{r}
df[, 3] <- as.factor(df[, 3])

str(df)
```

When you are done, use the `str()` command to show the data type for each variable in your dataframe.

# Peer Commentary Final Reflections

I thought you did a great job for this homework assignment! You took good notes (more than me for sure!) and you were able to solve the challenges without much issues! Your solutions were different to mine and I appreciate learning things this way. I left you comments for each challenge. Overall, I think that presentation, code organization, and code explanation could be made simpler to help people seeing your code for the first time understand your logic of solving things!
